An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-06 main.tex

\section{Applications of generalised Parikh images}\label{sec:applications}

\subsection{Example~1: String length}
\Fudge{We have already illustrated refer intro}
A useful example of such a simplifying morphism can express string length, the
problem that originally motivated our study of the Parikh map. We let $\Monoid =
\left(\mathbb{Z};+;0\right)$ and let $\Map(x) = 1$. Then the length of a string
$s = s_1 \RepeatSum{\Concat}  s_n$ is given by $\Map(s) = \sum \Map(s_i) = 1
\RepeatSum{+} 1$. Note that if we substitute vectors and vector addition for
this monoid, we obtain the regular Parikh map.

Intuitively, this often produces a simpler problem than computing the Parikh
image, as we now ignore not just the order of characters but also their
differences.

>>>>

\section{Generalised Parikh Maps}\label{sec:generalised}

It is easy to see that the Parikh map~$\ParikhMap$ represents a
homomorphism from the (free) non-commutative monoid~$\Strings$ to the
(free) commutative monoid~$\Naturals^k$. As we are often not
interested in the full Parikh image, but only in some projections of
it, for the purposes of this paper we use a generalised version of the
Parikh map. We consider \emph{arbitrary} homomorphisms
$\Map:\: \Sigma^* \to \Monoid$, where
$\Monoid = \left(X;+_{\Monoid};0_{\Monoid}\right)$ is a commutative
monoid; this means that $\Map(\epsilon) = 0_{\Monoid}$ and
$\Map(u \Concat w) = \Map(u) +_{\Monoid} \Map(w)$. We give several
examples of such generalised Parikh image later in this section.

Observe that every homomorphism~$\Map:\: \Sigma^* \to \Monoid$ can be
represented as the composition~$h' \circ \ParikhMap$, for some
homomorphism~$h' : \Naturals^k \to M$. One of the insights underlying our
approach is that it is more efficient to directly compute a
generalised Parikh image~$\Map(\Language)$, than to first compute the
standard image~$\ParikhMap(\Language)$ followed by projection to
some property of interest.

%\subsection{Applications of generalised Parikh images}\label{sec:applications}

\subsection{Example~1: Reasoning about string length}

A useful example of such a simplifying morphism can express string
length, the problem that originally motivated our study of the Parikh
map. We let $\Monoid = \left(\mathbb{Z};+;0\right)$, and define the
homomorphism~$\Map$ by $\Map(a) = 1$ for all
characters~$a \in \Alphabet$. Then the length of a string
$s = s_1 \RepeatSum{\Concat} s_n$ is given by
$\Map(s) = \sum \Map(s_i) = 1 \RepeatSum{+} 1 = n$.

Intuitively, this produces a simpler problem than computing the Parikh image, as
we now ignore not just the order of characters but also their differences.
\iffalse
This
means that the branching automaton of \cref{fig:len-branch} becomes
equivalent to the non-branching automaton of \cref{fig:len-branch-free}.
Note that the elimination of the choice of paths also effectively eliminates the
possibility of disconnecting one of the self-loops. This, in turn, means that we
can eliminate all of the connectedness implications of the quantified formula,
which would in turn simplify it (and the quantifier elimination of it)
considerably.
\fi
